# Partial Derivatives

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
# Load packages
library(mosaic)
library(mosaicCore)
library(mosaicCalc)
library(latticeExtra)
```

## Rundown

::: {.callout-note title="Learning goals"}
-   Define what a partial derivative is.
-   Explain the difference between $f_x(x,y)$ and $f_y(x,y)$
-   Estimate the partial derivatives from a table of data
-   Estimate the partial derivatives from a contour plot.
-   Use desmos to approximate the partial derivatives
:::

## Activities

### Estimating From Data

The maximum duration of a scuba dive $T(V,D)$ (in minutes) depends on the volume $V$ of air (at sea-level pressure) in the tank and the depth $D$ (in feet) of the dive. The following table shows the estimated dive times for various combinations of air volume and dive depth.

+:------------------------:+:---------------:+
| **Dive Time (minutes)**  | **Air Volume**  |
+--------------------------+-----------------+
| **20 ft^3^**             | **40 ft^3^**    |
+--------------------------+-----------------+
| **Depth**                | **20 ft**       |
+--------------------------+-----------------+
| **40 ft**                | 9.0             |
+--------------------------+-----------------+
| **60 ft**                | 7.1             |
+--------------------------+-----------------+
| **80 ft**                | 5.8             |
+--------------------------+-----------------+

Use this table to estimate the derivatives $\displaystyle{\frac{\partial T}{\partial V}}$ and $\displaystyle{\frac{\partial T}{\partial D}}$ at the point $(V=60, T=40)$. Give units for each of these derivatives and then interpret what these numbers mean.

### Estimating From a Contour Plot

The production function $F(L,K)$ for a factory gives the number of units produced by labor $L$ and capital $K$ (both measured in dollars). This production function is given by $$
F(L,K) = 10 L^{2/7}K^{5/7}.
$$ Here is a contour plot of the production function.

```{r, echo=FALSE, fig.asp=1, include=FALSE}


plotFun(10* x^{2/7} * y^{5/7}~x&y, xlim=c(0,1000), ylim=c(0,1000), levels = seq(1000,9000,1000), filled = FALSE, xlab="labor L", ylab = "capital K", lwd = 3)
ladd( panel.abline(h=seq(100,900,100),v=seq(100,900,100),col='gray50') )
```

```{r, echo=FALSE, fig.asp=1}

ladd( panel.abline(h=seq(100,900,100),v=seq(100,900,100),col='gray50') )
```

1.  Estimate the production output for $L=400$ and $K=600$.
2.  Estimate the partial derivative $\displaystyle{\frac{\partial F}{\partial L}}$ at $(400,600)$.
3.  Estimate the partial derivative $\displaystyle{\frac{\partial F}{\partial K}}$ at $(400,600)$.

### Estimating Using Desmos

Use desmos to define the prodution function $F(L,K) = L^{2/7}  K^{5/7}$ from the previous problem and to answer the analogous questions

1.  Calculate the production outpuut $F(400,600)$.
2.  Estimate the partial derivative $\displaystyle{\frac{\partial F}{\partial L}}$ at $(400,600)$. Your answer must be correct to two decimal places.
3.  Estimate the partial derivative $\displaystyle{\frac{\partial F}{\partial K}}$ at $(400,600)$. Your answer must be correct to two decimal places.

Compare your three answers to the estimates from the contour plot in the previous problem.

\
\
\
\

## Solutions

<details>

<summary>Click for Solutions</summary>

### Estimating From Data

We estimate the partial derivatives. We have $$
T_V(60,40) \approx \frac{T(80,40) - T(60,40)}{80-60}
= \frac{36 - 27}{80-60} = \frac{9}{20} = 0.45 \mbox{ min/ft}^3
$$ where we have used the interval $60 \leq V \leq 80$ for our estimate. We could also use $40 \leq V \leq 80$ or $40 \leq V \leq 80$ as the interval for our estimate. $$
T_D(60,40) \approx \frac{T(60,60) - T(60,40)}{60-40}
= \frac{21.3 - 27}{60-40} = - \frac{5.7}{20} = -0.285 \mbox{ min/ft}
$$ where we have used the interval $40 \leq D \leq 60$ for our estimate. We could also use $20 \leq D \leq 40$ or $20 \leq D \leq 60$ as the interval for our estimate.

The units for $T_V(V,D)$ are minutes per cubic feet. An additional cubic foot of air will allow you to dive for an extra 0.45 minutes.

The units for $T_V(V,D)$ are minutes per feet. An additional foot of dive depth will reduce your dive time: the change is $-0.285$ minutes.

### Estimating From a Contour Plot

We have $$
F(400,600) \approx 5300.
$$

Let's estimate by using the contour line $F(L,K) = 5000$. I estimate that $F(325,600)=5000$ and that $F(400,550)=5000$. Therefore

$$
F_L(400,600) \approx \frac{5300 - 5000}{400 - 325} = \frac{300}{75} = 4
$$ and $$
F_K(400,600) \approx \frac{5300 - 5000}{600-550}
= \frac{300}{50} = 6.
$$

### Estimating Using Desmos

Here is a [link to my desmos calculations](https://www.desmos.com/calculator/piexczmrsn).

1.  $F(400,600) = 534.37$
2.  $F_L(400,600) = 0.38$
3.  $F(400,600) = 0.64$

</details>
